shorts: An R Package for Modeling Short Sprints

### 2021-04-14 paper.Rmd
Reduced version for the PeerJ

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  fig.retina = 0.8, # figures are either vectors or 300 dpi diagrams
  dpi = 300,
  out.width = "100%",
  fig.align = "center",
  fig.width = 6,
  fig.height = 6 * 0.618, # 1 / phi
  fig.show = "hold",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  width = 70
)

# automatically create a bib database for R packages
#knitr::write_bib(c(
#  .packages(), "bookdown", "knitr", "rmarkdown", "shorts", "nlme", "LambertW"
#), "packages.bib")

@@

options(
  digits = 3,
  "width" = 65,
  scipen = 999)
```

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (<5 s) of maximal efforts [@mangineSpeedForcePower2014], whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) [@ward-smithEnergyConversionStrategies2001], but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)[@brownAssessmentLinearSprinting2004]. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. Radar and laser technology are frequently used laboratory-grade methods [@buchheitMechanicalDeterminantsAcceleration2014; @edwardsSprintAccelerationCharacteristics2020; @jimenez-reyesRelationshipVerticalHorizontal2018; @marcote-pequenoAssociationForceVelocity2019] but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The **shorts** package [@R-shorts], written in the R language [@R-base], represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016].

@@

gg <- ggplot(
  kinematics_long,
  aes(x = time, y = value, color = Athlete)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(
    ~variable,
    scales = "free_y",
    labeller = label_parsed
  ) +
  ylab(NULL) +
  xlab("Time (s)") +
  theme(
    legend.position = "top",
    legend.title = element_blank())

@@

Plotting acceleration against velocity (Figure \@ref(fig:four-athletes-profile)), we will get *Acceleration-Velocity Profile*, which is linear, according to the mathematical model. If the athlete's body mass (kg) is known, as well as additional air resistance parameters (see [Air resistance and the calculation of force and mechanical power] section of this paper), *Force-Velocity Profile* can be estimated (see [Force-Velocity profile] section of this paper).

```{r four-athletes-profile, echo=FALSE, fig.cap="Acceleration-Velocity profile of four athletes with different MSS and MAC parameters."}
gg <- ggplot(
  kinematics,
  aes(x = velocity, y = acceleration, color = Athlete)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  ylab(expression("Acceleration (ms"^-2 * ")")) +
  xlab(expression("Velocity (ms"^-1 * ")")) +
  theme(
    legend.position = "top",
    legend.title = element_blank())

@@

Besides providing *residual standard error* (RSE), **shorts** functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the `nls()` function:

```{r}
summary(m1)
```

To extract estimated model parameters, use S3 `coef()` method:

```{r}
coef(m1)
```

@@

```{r}
# Predict time at distance
predict_time_at_distance(
  distance = split_distance,
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU
)

# Predict acceleration at time
predict_acceleration_at_time(
  time = c(0, 1, 2, 3, 4, 5, 6),
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU
)
```

@@

```{r}
# To calculate horizontal force produced
predict_force_at_distance(
  distance = split_distance,
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = 80,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)

# To calculate power produced
predict_power_at_distance(
  distance = split_distance,
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = 80,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)
```

@@

```{r}
df <- predict_kinematics(
  m1,
  max_time = 6,
  frequency = 100,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = 80,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)

head(df)
```

@@

Additional information about mixed-effects model performed using the `nlme` package [@R-nlme] can be obtained using `summary()` function:

```{r}
summary(m2)
```

S3 method `coef()` when applied on mixed-model result will return both the fixed and random effects:

```{r}
coef(m2)
```

To create a simple plot of the model, use S3 `plot()` method:

@@

ggplot(
  filter(df, distance < 40),
  aes(x = distance, y = value, group = athlete, color = athlete)
) +
  theme_bw(8) +
  facet_wrap(~name, scales = "free_y") +
  geom_line(alpha = 0.7) +
  ylab(NULL) +
  xlab("Distance (m)") +
  theme(
    legend.position = "top",
    legend.title = element_blank())
```

@@

m3 <- model_using_radar(
  velocity = sprint_velocity,
  time = sprint_time
)

m3
```

```{r}
plot(m3) + theme_bw(8)
```

@@

```{r}
m3_weighted <- model_using_radar(
  velocity = sprint_velocity,
  time = sprint_time,
  weights = 1 / (sprint_velocity + 1)
)

m3_weighted
```

@@

m4 <- mixed_model_using_radar(
  radar_gun_data,
  time = "time",
  velocity = "velocity",
  athlete = "athlete"
)

m4
```

```{r}
plot(m4) + theme_bw(8)
```

@@

```{r}
plot(fvp, "time") + theme_bw(8)
```

@@

ggplot(
  plot_df,
  aes(x = value, y = velocity, color = `Sync issue`)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  xlab("Time (s)") +
  ylab(expression("Velocity (" * ms^-1 * ")")) +
  theme(
    legend.title = element_blank(), 
    legend.position = "top")
```

The following three models estimate MSS and TAU from the three datasets:

```{r}
# Without synchronization issues
m5 <- model_using_radar(
  velocity = df$velocity,
  time = df$`true time`
)

@@

rbind(
  data.frame(
    model = "True time",
    t(coef(m5))
  ),
  data.frame(
    model = "Added 0.5s time",
    t(coef(m6))
  ),
  data.frame(
    model = "Deducted 0.5s time",
    t(coef(m7))
  )
)
```

@@

m10

# Mixed model with time correction being random effect
m11 <- mixed_model_using_radar_with_time_correction(
  radar_gun_data,
  time = "time",
  velocity = "velocity",
  athlete = "athlete",
  random = MSS + TAU + time_correction ~ 1
)

m11
```

@@

John represents the *theoretical model*, in which we assume that the initial force production and the timing initiation are perfectly synchronized. Jack, on the other hand, represents a *practical model*, and decides to move slightly behind the initial timing gate (i.e. for 0.5 m) and use body rocking to initiate the sprint start. In other words, Jack is using a *flying start*, a common scenario when testing field sports athletes. Let's see how their sprint outcomes differ.

@@

  # Jack's performance
  jack_distance = distance + 0.5,
  jack_true_time = predict_time_at_distance(jack_distance, MSS, TAU),
  time_05m = predict_time_at_distance(0.5, MSS, TAU),
  jack_time = jack_true_time - time_05m
)

split_times
```

And here is a graphical representation of the sprint splits:

```{r}
plot_df <- split_times %>%
  select(distance, john_time, jack_time) %>%
  rename(John = john_time, Jack = jack_time) %>%
  pivot_longer(cols = -1, names_to = "athlete", values_to = "time") %>%
  mutate(distance = factor(distance))

ggplot(
  plot_df,
  aes(x = distance, y = time, color = athlete, group = athlete)
) +
  theme_bw(8) +
  geom_point() +
  geom_line() +
  xlab("Distance (m)") +
  ylab("Time (s)") +
  theme(
    legend.title = element_blank(), 
    legend.position = "top")
```

@@

sprint_parameters
```

@@

Below is a simulation sprint with 5, 10, 20, 30, 40, and 50 m splits, with MSS and MAC varying from 6 to 9 ($ms^{-1}$ and $ms^{-2}$ respectively), and flying start distance varying from 0 to 1 m.

```{r}
sim_df <- expand.grid(
  MSS = c(6, 7, 8, 9),
  MAC = c(6, 7, 8, 9),
  flying_start_distance = c(
    seq(0, 0.001, length.out = 20),
    seq(0.001, 0.01, length.out = 20),
    seq(0.01, 0.1, length.out = 20),
    seq(0.1, 1, length.out = 20)
  ),
  distance = c(5, 10, 20, 30, 40, 50)
)

sim_df <- sim_df %>%
  mutate(
    TAU = MSS / MAC,
    PMAX = MSS * MAC / 4,
    true_distance = distance + flying_start_distance,
    true_time = predict_time_at_distance(true_distance, MSS, TAU),
    stolen_time = predict_time_at_distance(
      flying_start_distance, MSS, TAU),
    time = true_time - stolen_time
  )

@@

Now when we have a simulation dataset, we can check the model estimates and predictions, given the flying start distance:

```{r}
# Prediction wrapper
pred_wrapper <- function(data) {
  model <- model_using_splits(
    distance = data$distance,
    time = data$time
  )

@@

```{r}
# Estimates plot
df <- model_df %>%
  group_by(MSS, TAU, flying_start_distance) %>%
  slice(1) %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2))
  )

# MSS
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MSS, color = MAC_string)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MSS (m/s)") +
  theme(
    legend.title = element_blank(), 
    legend.position = "top")
```

@@

```{r}
# MAC
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MAC, color = MSS_string)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~MAC_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MAC (m/s/s)") +
  theme(
    legend.title = element_blank(), 
    legend.position = "top")
```

@@

```{r}
# PMAX
ggplot(
  df,
  aes(x = flying_start_distance, y = est_PMAX, color = MSS_string)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~MAC_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated PMAX (W/kg)") +
  theme(
    legend.title = element_blank(), 
    legend.position = "top")
```

@@

```{r}
# Residuals
model_df <- model_df %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2)),
    group = paste(MSS, MAC, flying_start_distance)
  )

ggplot(
  model_df,
  aes(
    y = residuals,
    x = distance,
    color = flying_start_distance,
    group = group)
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  facet_grid(MSS_string ~ MAC_string) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") + 
  labs(color = "Flying start distance")
```

@@

```{r}
ggplot(
  model_df,
  aes(
    y = residuals,
    x = distance,
    color = flying_start_distance,
    group = group)
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") + 
  labs(color = "Flying start distance")
```

@@

jack_parameters
```

@@

We will get back to these issues later, but we can examine how these models perform using simulated data with varying flying start distance. The following code contains the wrapper that performs all four models (no correction, fixed time correction, estimated time correction, and estimated time and distance correction):

```{r}
pred_wrapper <- function(data) {
  no_correction <- model_using_splits(
    distance = data$distance,
    time = data$time
  )

@@

```{r}
model_df$model <- factor(
  model_df$model,
  levels = c(
    "No correction",
    "Fixed correction +0.3s",
    "Fixed correction +0.5s",
    "Time correction",
    "Time and distance correction"
  )
)
# Estimates plot
df <- model_df %>%
  group_by(MSS, TAU, flying_start_distance, model) %>%
  slice(1) %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2))
  )

# MSS
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MSS, color = model)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_grid(MSS_string ~ MAC_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MSS (m/s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top")
```

@@

```{r}
# MAC
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MAC, color = model)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_grid(MAC_string ~ MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MAC (m/s/s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top")
```

@@

```{r}
# PMAX
ggplot(
  df,
  aes(x = flying_start_distance, y = est_PMAX, color = model)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_grid(MAC_string ~ MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated PMAX (W/kg)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top")
```

@@

```{r}
# time_correction
ggplot(
  df,
  aes(
    x = flying_start_distance,
    y = est_time_correction,
    color = model)) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  geom_line(
    aes(y = stolen_time), color = "black", linetype = "dashed") +
  facet_grid(MAC_string ~ MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated time correction (s)")  +
  theme(
    legend.title = element_blank(),
    legend.position = "top")
```

The following figure depicts estimated distance correction, and same as with the time correction, only the time and distance correction model estimated the distance correction correctly (i.e., flying start distance; indicated by the dashed line on the figure, which represents *identity line* since flying start distance is already on the x-axis).

```{r}
# distance_correction
ggplot(
  df,
  aes(
    x = flying_start_distance,
    y = est_distance_correction, 
    color = model)) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  geom_abline(slope = 1, color = "black", linetype = "dashed") +
  facet_grid(MAC_string ~ MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated distance correction (m)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top")
```

@@

# Add small noise to allow model fit
set.seed(1667)
rand_noise <- rnorm(nrow(sim_df), 0, 10^-4)
sim_df$time <- sim_df$time + rand_noise
```

>>>>

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  fig.retina = 0.8, # figures are either vectors or 300 dpi diagrams
  dpi = 300,
  out.width = "90%",
  fig.align = "center",
  fig.width = 5,
  fig.height = 5 * 0.618, # 1 / phi
  fig.show = "hold",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  width = 65
)

# automatically create a bib database for R packages
# knitr::write_bib(c(
#  .packages(), "bookdown", "knitr", "rmarkdown", "shorts", "nlme", "LambertW"
# ), "packages.bib")

@@

options(
  digits = 3,
  "width" = 65,
  scipen = 999
)
```

@@

Short sprint performance is one of the most distinguishable and admired physical traits in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (<5 s) of maximal efforts [@mangineSpeedForcePower2014], whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) [@ward-smithEnergyConversionStrategies2001], but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)[@brownAssessmentLinearSprinting2004]. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. Radar and laser technology are frequently used laboratory-grade methods [@buchheitMechanicalDeterminantsAcceleration2014; @edwardsSprintAccelerationCharacteristics2020; @jimenez-reyesRelationshipVerticalHorizontal2018; @marcote-pequenoAssociationForceVelocity2019] but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance.

The **shorts** package [@R-shorts], written in the R language [@R-base], represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. To best of our knowledge, scientist, researchers, and coaches have been performing short sprints modeling using the built-in solver function of Excel (Microsoft Corporation, Redmond, Washington, United States) [@morinSpreadsheetSprintAcceleration2017; @morinSpreadsheetSprintAcceleration2019; @stenrothForcevelocityProfilingIce2020; @stenrothSpreadsheetSprintAcceleration2020; @samozinoSimpleMethodMeasuring2016; @clarkNFLCombine40Yard2017; @morinSimpleMethodComputing2019], which makes the **shorts** package major improvement in ease-of-use, speed, transparency, reproducibility, and more feature-rich model fitting.

@@

gg <- ggplot(
  kinematics_long,
  aes(x = time, y = value, color = Athlete)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(
    ~variable,
    scales = "free_y",
    labeller = label_parsed
  ) +
  ylab(NULL) +
  xlab("Time (s)") +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )

@@

Plotting acceleration against velocity (Figure \@ref(fig:four-athletes-profile)), we will get *Acceleration-Velocity Profile*, which is linear, according to the mathematical model. If the athlete's body mass (kg) is known, as well as additional air resistance parameters (see [Air resistance and the calculation of force and mechanical power] section of this paper), *Force-Velocity Profile* can be estimated (see [Force-Velocity profile] section of this paper). 
```{r four-athletes-profile, echo=FALSE, fig.cap="Acceleration-Velocity profile of four athletes with different MSS and MAC parameters."}
gg <- ggplot(
  kinematics,
  aes(x = velocity, y = acceleration, color = Athlete)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  ylab(expression("Acceleration (ms"^-2 * ")")) +
  xlab(expression("Velocity (ms"^-1 * ")")) +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )

@@

Besides providing *residual standard error* (RSE), **shorts** functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the `nls()` function (i.e., by using the S3 `summary()` method). To extract estimated model parameters, use S3 `coef()` method.

@@

```{r}
# Predict time at distance
predict_time_at_distance(
  distance = split_distance,
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU
)
```

@@

```{r}
# To calculate horizontal force produced
predict_force_at_distance(
  distance = split_distance,
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = 80,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)
```

@@

```{r}
df <- predict_kinematics(
  m1,
  max_time = 6,
  frequency = 100,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = 80,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)
```

@@

Additional information about mixed-effects model performed using the `nlme` package [@R-nlme] can be obtained using `summary()` function. S3 method `coef()` when applied on mixed-model result will return both the fixed and random effects. To create a simple plot of the model, use S3 `plot()` method.

```{r include=FALSE}
summary(m2)
coef(m2)
```

@@

ggplot(
  filter(df, distance < 40),
  aes(x = distance, y = value, group = athlete, color = athlete)
) +
  theme_bw(8) +
  facet_wrap(~name, scales = "free_y") +
  geom_line(alpha = 0.7) +
  ylab(NULL) +
  xlab("Distance (m)") +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )
```

@@

m3 <- model_using_radar(
  velocity = sprint_velocity,
  time = sprint_time
)
```

@@

```{r}
m3_weighted <- model_using_radar(
  velocity = sprint_velocity,
  time = sprint_time,
  weights = 1 / (sprint_velocity + 1)
)
```

@@

m4 <- mixed_model_using_radar(
  radar_gun_data,
  time = "time",
  velocity = "velocity",
  athlete = "athlete"
)
```

@@

```{r results='hide', fig.show='hide'}
plot(fvp, "time") + theme_bw(8)
```

@@

ggplot(
  plot_df,
  aes(x = value, y = velocity, color = `Sync issue`)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  xlab("Time (s)") +
  ylab(expression("Velocity (" * ms^-1 * ")")) +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

The following three models estimate MSS and TAU from the three datasets. Tables are created using the `kable()` function from the **knitr** package [@knitr2014; @knitr2015; @R-knitr] and `kable_styling()` function from the **kableExtra** package [@R-kableExtra].

```{r}
require(knitr)
require(kableExtra)

# Without synchronization issues
m5 <- model_using_radar(
  velocity = df$velocity,
  time = df$`true time`
)

@@

model_df <- rbind(
  data.frame(
    model = "True time",
    t(coef(m5))
  ),
  data.frame(
    model = "Added 0.5s time",
    t(coef(m6))
  ),
  data.frame(
    model = "Deducted 0.5s time",
    t(coef(m7))
  )
)

kable(model_df, digits = 2, booktabs = TRUE)
```

@@

# Mixed model with time correction being random effect
m11 <- mixed_model_using_radar_with_time_correction(
  radar_gun_data,
  time = "time",
  velocity = "velocity",
  athlete = "athlete",
  random = MSS + TAU + time_correction ~ 1
)
```

@@

John represents the *theoretical model*, in which we assume that the initial force production and the timing initiation are perfectly synchronized. Jack, on the other hand, represents a *practical model*, and decides to move slightly behind the initial timing gate (i.e. for 0.5 m) and use body rocking to initiate the sprint start. In other words, Jack is using a *flying start*, a common scenario when testing field sports athletes.

@@

  # Jack's performance
  jack_distance = distance + 0.5,
  jack_true_time = predict_time_at_distance(jack_distance, MSS, TAU),
  time_05m = predict_time_at_distance(0.5, MSS, TAU),
  jack_time = jack_true_time - time_05m
)
```

Here is a graphical representation of the sprint splits (please refer to the Supplemental files for the R code):

```{r echo=FALSE}
plot_df <- split_times %>%
  select(distance, john_time, jack_time) %>%
  rename(John = john_time, Jack = jack_time) %>%
  pivot_longer(cols = -1, names_to = "athlete", values_to = "time") %>%
  mutate(distance = factor(distance))

ggplot(
  plot_df,
  aes(x = distance, y = time, color = athlete, group = athlete)
) +
  theme_bw(8) +
  geom_point() +
  geom_line() +
  xlab("Distance (m)") +
  ylab("Time (s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

@@

kable(sprint_parameters, digits = 2, booktabs = TRUE)
```

@@

To explore this further, we have run a simple simulation with 5, 10, 20, 30, 40, and 50 m splits, with MSS and MAC varying from 6 to 9 ($ms^{-1}$ and $ms^{-2}$ respectively), and flying start distance varying from 0 to 1 m (please refer to the Supplemental files for the R code). With a simulation dataset, we can check the model estimates and predictions, given the flying start distance.

```{r include=FALSE}
sim_df <- expand.grid(
  MSS = c(6, 7, 8, 9),
  MAC = c(6, 7, 8, 9),
  flying_start_distance = c(
    seq(0, 0.001, length.out = 20),
    seq(0.001, 0.01, length.out = 20),
    seq(0.01, 0.1, length.out = 20),
    seq(0.1, 1, length.out = 20)
  ),
  distance = c(5, 10, 20, 30, 40, 50)
)

sim_df <- sim_df %>%
  mutate(
    TAU = MSS / MAC,
    PMAX = MSS * MAC / 4,
    true_distance = distance + flying_start_distance,
    true_time = predict_time_at_distance(true_distance, MSS, TAU),
    stolen_time = predict_time_at_distance(
      flying_start_distance, MSS, TAU
    ),
    time = true_time - stolen_time
  )

@@

```{r include=FALSE}
# Prediction wrapper
pred_wrapper <- function(data) {
  model <- model_using_splits(
    distance = data$distance,
    time = data$time
  )

@@

```{r echo=FALSE}
# Estimates plot
df <- model_df %>%
  group_by(MSS, TAU, flying_start_distance) %>%
  slice(1) %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2))
  )

# MSS
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MSS, color = MAC_string)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MSS (m/s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

@@

```{r echo=FALSE}
# MAC
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MAC, color = MSS_string)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~MAC_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MAC (m/s/s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

@@

```{r echo=FALSE}
# PMAX
ggplot(
  df,
  aes(x = flying_start_distance, y = est_PMAX, color = MSS_string)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~MAC_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated PMAX (W/kg)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

@@

```{r echo=FALSE}
# Residuals
model_df <- model_df %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2)),
    group = paste(MSS, MAC, flying_start_distance)
  )

ggplot(
  model_df,
  aes(
    y = residuals,
    x = distance,
    color = flying_start_distance,
    group = group
  )
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  facet_grid(MSS_string ~ MAC_string) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") +
  labs(color = "Flying start distance")
```

@@

```{r echo=FALSE}
ggplot(
  model_df,
  aes(
    y = residuals,
    x = distance,
    color = flying_start_distance,
    group = group
  )
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") +
  labs(color = "Flying start distance")
```

@@

kable(jack_parameters, digits = 2, booktabs = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"))
```

@@

We will get back to these issues later, but we can examine how these models (no correction, fixed time correction, estimated time correction, and estimated time and distance correction) perform using simulated data with varying flying start distance (please refer to the Supplemental files for the R code).

```{r echo=FALSE}
pred_wrapper <- function(data) {
  no_correction <- model_using_splits(
    distance = data$distance,
    time = data$time
  )

@@

```{r echo=FALSE}
model_df$model <- factor(
  model_df$model,
  levels = c(
    "No correction",
    "Fixed correction +0.3s",
    "Fixed correction +0.5s",
    "Time correction",
    "Time and distance correction"
  )
)
# Estimates plot
df <- model_df %>%
  group_by(MSS, TAU, flying_start_distance, model) %>%
  slice(1) %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2))
  )

# MSS
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MSS, color = model)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_grid(MSS_string ~ MAC_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MSS (m/s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

@@

```{r echo=FALSE}
# MAC
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MAC, color = model)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_grid(MAC_string ~ MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MAC (m/s/s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

@@

```{r echo=FALSE}
# PMAX
ggplot(
  df,
  aes(x = flying_start_distance, y = est_PMAX, color = model)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_grid(MAC_string ~ MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated PMAX (W/kg)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

@@

```{r echo=FALSE}
# time_correction
ggplot(
  df,
  aes(
    x = flying_start_distance,
    y = est_time_correction,
    color = model
  )
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  geom_line(
    aes(y = stolen_time),
    color = "black", linetype = "dashed"
  ) +
  facet_grid(MAC_string ~ MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated time correction (s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

The next figure depicts estimated distance correction, and same as with the time correction, only the time and distance correction model estimated the distance correction correctly (i.e., flying start distance; indicated by the dashed line on the figure, which represents *identity line* since flying start distance is already on the x-axis).

```{r echo=FALSE}
# distance_correction
ggplot(
  df,
  aes(
    x = flying_start_distance,
    y = est_distance_correction,
    color = model
  )
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  geom_abline(slope = 1, color = "black", linetype = "dashed") +
  facet_grid(MAC_string ~ MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated distance correction (m)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

@@

# Add small noise to allow model fit
set.seed(1667)
rand_noise <- rnorm(nrow(sim_df), 0, 10^-4)
sim_df$time <- sim_df$time + rand_noise
